\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-05-06 honors_thesis/journal_paper/paper.tex
Done with my final pass through the journal paper

Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
the standard PSO to make better use of the information obtained by the
speculative evaluations.  Using these algorithmic modifications, we see
dramatic improvements in the effectiveness of the algorithm.

@@

In this paper we consider PSO parallelization strategies for clusters of
thousands of processors and functions for which a single evaluation will take
at least tens of seconds, but probably minutes or perhaps hours.  Our purpose
is to explore the question of what to do with a thousand processors when 50 or
100 particles is the recommended swarm size, and simply increasing the swarm
size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction) techniques commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g. a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes
can be undone.  If the processor guesses right, execution is
much farther ahead than if it had idly waited on the memory reference.  If
it guesses wrong, execution restarts where it would have been anyway.

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once.  In
\sec{relax}, we discuss various methods of improving the performance of
speculative evaluation in PSO, all of which break the requirement of strictly
reproducing the behavior of the original algorithm.  \sec{results} presents the
results of using our methods in comparison with standard PSO.  In
\sec{conclusion} and \sec{future} we conclude and discuss future work.

@@

The PSO algorithm can be trivially parallelized by distributing the processing
needed for each particle on across processors.  But as we have seen in
\fig{evals-sphere}, for some functions, and for large numbers of
processors, just adding particles reaches a point of diminishing returns.
That is, adding processors does not help us reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
evaluation approach that allows us to perform iterations two at a time.

@@

The refactoring hinges on the idea that once the random coefficients are
determined, there are only a few possibile updates to $\nbest$ and $\pbest$.
For the Ring topology there are 7 possible update cases, identified in
Table~\ref{tab:evals}.  We label each case with an identifier referring to the
source of the update: a minus sign ($-$) represents no update, $L$ represents
an update to $\nbest$ coming from the left neighbor, $R$ represents an update
to $\nbest$ coming from the right neighbor, and $S$ represents an update to
either $\pbest$ or $\nbest$ coming from the particle itself.  As an example,
$\casexn$ refers to the case that the particle finds a new personal best, but
neither it nor its neighbors found a position that updated its neighborhood
best.  In the equations that follow, we refer to an unspecified update case as
$\casegen$, and to the set of cases collectively as $\caseset$.

@@

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \wedge \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

@@

In this form the important point to notice is that there are only 7 values (for
this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$ (note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration).  Thus PSO has been refactored
such that the algorithm can begin computing all 7 of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative algorithm computes
$\ofunc(\specpos_{t+1}^{\casegen})$ for all $\casegen \in \caseset$, even those
for which $\indic_{t+1}^{\casegen} = 0$, these extra computations will be
ignored, and might just as well never have been computed.  We call the set of
computations $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
``speculative children'' because only one of them will be needed.

To see the value of this refactoring, suppose that 1000 processors are
available, and that the evaluation of the function to be optimized takes one
hour.  If we only want a swarm of 100 particles, 900 of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $\ofunc(\specpos_{t+1}^{c})$ to be computed at the same time as
$\ofunc(\ppos_{t})$, we would use 800 of the 1000 processors and perform two
iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If we were not performing speculative
evaluation, those processors might very well be used to add particles to the
swarm.  This then raises the question of whether a swarm of 100 particles doing
twice as many iterations outperforms a swarm of 800 particles.  We show in the
\sec{results} that in many, though not all, instances, a smaller swarm
performing more iterations does in fact outperform a larger swarm.

@@

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  We can try to make use of those evaluations instead of
throwing them away.

To use the extra speculative evaluations, instead of choosing the speculative
child that matches the branch that the original PSO would have taken, we take
the child that has the best value.  The methodology is exactly the same as
above except for the process of choosing which speculative child to accept.
The only change needed in \alg{centralized} is in step 7, where
$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
instead of with the matching branch.

This can be thought of as drawing a number of samples from the next iteration
and accepting the best one.  Speculative particles that move in good directions
are kept.  It seems this modification to PSO would produce more exploitation,
decreasing the amount of exploration.  With functions that are less deceptive
we would expect this method to work well, while performance on more deceptive
functions would probably be hurt.

@@

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results for only one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

@@

There are also many ways to speculate several iterations ahead, and we again
only show results for one of them.  The combination of branches we tried uses
seven speculative evaluations per particle, matching the swarm size of the
original speculative algorithm.  The seven evaluations we used corresponded to
several iterations of branches \casepn\ and \casexn.  Two of the evaluations
were just one iteration ahead, four were two iterations ahead, and one was
three iterations ahead.  The evaluations that were one iteration ahead were
branches \casepn\ and \casexn; those that went two iterations ahead were formed
by taking either branch \casepn\ or \casexn and then branch \casepn\ or
\casexn; and the evaluation going three iterations ahead followed branch
\casepn\ on all three iterations.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.}
  \label{fig:rastrigin}
\end{figure}

@@

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  We expect our methods to be perfect
for such functions, and our results show this intuition again to be correct.
First we compare the basic SEPSO algorithm to standard PSO.  In
\fig{basic-sphere} we can see that when both SEPSO and PSO are using a Random
topology, SEPSO clearly outperforms PSO.  However, the comparison between SEPSO
and PSO Complete is not as encouraging---PSO Complete performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:basic-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:sphere-pickbest}
\end{figure}

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iters by far outperforms every other algorithm on this function.  Many Iters on
average finds values that are more than 40 orders of magnitude better than
standard PSO using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel1}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:schwefel1}
\end{figure}

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than PSO Complete, with Many Iterations performing comparably to
PSO.  The only algorithm that performed better than PSO was Pick Best Pruned,
likely because of the larger swarm size available to it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel2}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations, comparing all speculative algorithms.}
  \label{fig:schwefel2}
\end{figure}

@@

Next we look at the function Griewank.  It is generally recommended to use the
Ring topology when optimizing the Griewank function, as Complete is prone to
premature convergence on a local
optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank has a global
optimum with a value of 0, and sometimes the swarm finds the optimum and
sometimes it doesn't.  Instead of showing average function value at each
iteration, a more informative plot for Griewank shows the percent of runs that
have found the global optimum by each iteration.  We ran 50 trials of each
experiment with Griewank, so that the curves are more smooth.  We show results
in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO algorithm finds the optimum faster
than the original PSO.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:griewank-pruned}
\end{figure}

With Griewank, the premature convergence problems inherent in picking the best
child are exacerbated when speculating several iterations ahead.  When Many
Iterations finds the optimum, it finds it quicker than any other method we
tried, on average four times faster than standard PSO.  However, it also gets
stuck and fails to find the optimum more than any other method.  The results
are shown in \fig{griewank-manyiters}. (TODO: Rerun Many Iterations with the
right swarm size)

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.}
  \label{fig:griewank-800}
\end{figure}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes doesn't.  Since we have already
demonstrated the convergence properties of the various algorithms with
Griewank, for Bohachevsky we merely present the best results.  In
\fig{bohachevsky} we can see that all of the speculative algorithms found the
optimum much quicker than PSO with a Random topology.  However, SEPSO was
slower than PSO Complete and got stuck around 15\% of the time.  Using Pick
Best, Pick Best Pruned, and Many Iterations all outperformed PSO Complete,
though Many Iterations only almost reached a 100\% success rate.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with a swarm that uses 800 processors per round
  of evaluations.}
  \label{fig:bohachevsky}
\end{figure}

>>>>

Speculative Evaluation in PSO can be enhanced by diverging from the behavior of
the standard PSO.  We present several variations on our speculative algorithm,
such as making use of the information obtained by the extra speculative
evaluations, and speculating more than one iteration ahead.  Using these
algorithmic modifications, we see dramatic improvements in the effectiveness of
the algorithm.

@@

In this paper we consider PSO parallelization strategies for clusters of
hundreds or thousands of processors and functions for which a single evaluation
will take at least tens of seconds, but probably minutes or perhaps hours.  Our
purpose is to explore the question of what to do with a thousand processors
when 50 or 100 particles is the recommended swarm size, and simply increasing
the swarm size encounters diminishing returns.

To solve this problem, we propose a method for performing two iterations of PSO
at the same time in parallel that we call speculative evaluation.  The name
comes from an analogy to speculative execution (also known as branch
prediction), a technique commonly used in processors.  Modern processors, when
faced with a branch on which they must wait (e.g., a memory cache miss), guess
which way the branch will go and start executing, ensuring that any changes can
be undone.  If the processor guesses right, execution is much farther ahead
than if it had idly waited on the memory reference.  If it guesses wrong,
execution restarts where it would have been anyway.

@@

The balance of this paper is organized as follows. \sec{pso} describes the
particle swarm optimization algorithm.  \sec{sepso} describes how speculative
evaluation can be done in parallel PSO to perform two iterations at once.  In
\sec{relax}, we discuss various methods of improving the performance of
speculative evaluation in PSO, all of which break the requirement of strictly
reproducing the behavior of the original algorithm.  \sec{setup} describes the
experiments we ran, and \sec{results} presents the results of those
experiments.  In \sec{conclusion} and \sec{future} we conclude and discuss
future work.

@@

The PSO algorithm can be trivially parallelized by distributing the computation
needed for each particle on across processors.  But as we have seen in
\fig{evals-sphere}, for some functions, and for large numbers of processors,
just adding particles reaches a point of diminishing returns.  That is, adding
processors does not help the algorithm reach any given level of fitness
appreciably faster.  Instead of adding particles we employ a speculative
approach that allows us to perform iterations two at a time.

@@

The refactoring hinges on the idea that once the random coefficients
$\pURand_{t}$ and $\pURand_{t}$ are determined, there are only a few possible
updates to $\nbest$ and $\pbest$.  For the Ring topology there are 7 possible
update cases, identified in Table~\ref{tab:evals}.  We label each case with an
identifier referring to the source of the update: a minus sign ($-$) represents
no update, $L$ represents an update to $\nbest$ coming from the left neighbor,
$R$ represents an update to $\nbest$ coming from the right neighbor, and $S$
represents an update to either $\pbest$ or $\nbest$ coming from the particle
itself.  As an example, $\casexn$ refers to the case that the particle finds a
new personal best, but neither it nor its neighbors find a position that
updated its neighborhood best.  In the equations that follow, we refer to an
unspecified update case as $\casegen$, and to the set of cases collectively as
$\caseset$.

@@

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

@@

In this form the important point to notice is that there are only $7$ values
(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$.  Note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration.  Thus PSO has been refactored such
that the algorithm can begin computing all $7$ of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
= 0$, these extra computations will be ignored, and might just as well never
have been computed.  We call the set of computations
$\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
children'' because only one of them will be needed.

To see the value of this refactoring, suppose that $1000$ processors are
available, and that the evaluation of the objective function takes one hour.
If we only want a swarm of $100$ particles, $900$ of the processors would be
sitting idle for an hour at every iteration, and it would take two hours to run
two iterations.  If instead we perform speculative evaluation, sending each of
the $7$ evaluations in $\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$
to be computed at the same time as $\ofunc(\ppos_{t})$, we would use $800$ of
the $1000$ processors and perform two iterations in one hour.

In order to do two iterations at once, we must use 8 times as many processors
as there are particles in the swarm.  If these processors were not performing
speculative evaluation, they might instead be used for function evaluations
needed to support a larger swarm.  This raises the question of whether a swarm
of 100 particles doing twice as many iterations outperforms a swarm of 800
particles.  We show in \sec{results} that in many, though not all, instances, a
smaller swarm performing more iterations does in fact outperform a larger
swarm.

@@

In performing speculative evaluation as we have described it, $2n+1$
speculative evaluations are done per particle, while all but one of them are
completely ignored.  It seems reasonable to try to make use of the information
obtained through those evaluations instead of ignoring it.

To make better use of the extra speculative evaluations, instead of choosing
the speculative child that matches the branch that the original PSO would have
taken, we take the child that has the best value.  The methodology is exactly
the same as above except for the process of choosing which speculative child to
accept.  The only change needed in \alg{centralized} is in step 7, where
$\noeval{\s}_{t+1}$ with the best value is chosen from $\noeval{\sset}_{t+1}$
instead of with the matching branch.

This can be thought of as drawing a number of samples from the next iteration
and accepting the best one.  Speculative particles that move in good directions
are kept.  It seems this modification to PSO would produce more exploitation,
decreasing the amount of exploration.  With functions that are less deceptive
we would expect this method to work well, while performance on more deceptive
functions would probably be hurt.  We call this technique Pick Best.

@@

The algorithms we compare are the standard PSO algorithm, SEPSO, and four
variations on SEPSO.  In presenting our results, we call the variant developed
in \sec{pickbest} Pick Best.  The algorithms described in \sec{pruning} through
\sec{wrong} are called Pick Best Pruned and Social Promotion Pruned, and the
algorithm in \sec{manyiters} we call Many Iterations.

When performing pruning in SEPSO in Pick Best Pruned and Social Promotion
Pruned, there are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results for only one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

@@

There are also many ways to speculate several iterations ahead, and with Many
Iterations we again only show results for one of them.  The combination of
branches we tried uses seven speculative evaluations per particle, matching the
swarm size of the original speculative algorithm.  The seven evaluations we
used corresponded to several iterations of branches \casepn\ and \casexn.  Two
of the evaluations were just one iteration ahead, four were two iterations
ahead, and one was three iterations ahead.  The evaluations that were one
iteration ahead were branches \casepn\ and \casexn; those that went two
iterations ahead were formed by taking either branch \casepn\ or \casexn and
then branch \casepn\ or \casexn; and the evaluation going three iterations
ahead followed branch \casepn\ on all three iterations.

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{rastrigin}
  \caption{Function Rastrigin with a swarm that uses 240 processors per round
  of evaluations.  All speculative algorithms use the Random topology.}
  \label{fig:rastrigin}
\end{figure}

@@

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima (only a single global optimum) and is best optimized in
terms of function evaluations with a small swarm using a Complete topology.  We
expect our methods to be perfect for such functions, and our results show this
intuition again to be correct.  First we compare the basic SEPSO algorithm to
standard PSO.  In \fig{basic-sphere} we can see that when both SEPSO and PSO
are using a Random topology, SEPSO clearly outperforms PSO.  However, the
comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  SEPSO uses the Random topology.}
  \label{fig:basic-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere2}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.  All speculative
  algorithms use the Random topology.}
  \label{fig:sphere-pickbest}
\end{figure}

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iterations by far outperforms every other algorithm on this function.  Many
Iterations on average finds values that are more than 40 orders of magnitude
better than standard PSO using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.  All speculative algorithms use the Random topology.}
  \label{fig:sphere-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel1}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations.  SEPSO uses a Random topology.}
  \label{fig:schwefel1}
\end{figure}

As can be seen in \fig{schwefel2}, almost all of the speculative algorithms
perform worse than PSO Complete, with Many Iterations performing comparably to
PSO.  However, one of the speculative algorithms did perform better than PSO,
Pick Best Pruned.  It seems that the larger swarm size available because of
pruning was the cause of the improved performance.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{schwefel2}
  \caption{Function Schwefel with a swarm that uses 240 processors per round of
  evaluations, comparing all speculative algorithms.  All speculative
  algorithms use the Random topology.}
  \label{fig:schwefel2}
\end{figure}

@@

Next we look at the function Griewank.  It is generally recommended to use the
Ring topology when optimizing the Griewank function, as Complete is prone to
premature convergence on a local
optimum~\cite{bratton-2007-defining-a-standard-for-pso}.  Griewank has a global
optimum with a value of 0, and sometimes the swarm finds the optimum and
sometimes it does not.  Instead of showing average function value at each
iteration, a more informative plot for Griewank shows the percent of runs that
have found the global optimum by each iteration.  We ran 50 trials of each
experiment with Griewank, so that the curves are more smooth.  We show results
in \fig{basic-griewank1} for swarms of size 30 and 240 using the Ring topology.
One can see in the figure that the SEPSO algorithm finds the optimum faster
than the original PSO.  However, because the swarm size is so small, SEPSO gets
stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank1}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations.  Instead of showing average function value, we show the
  percentage of runs that have found the global optimum by each iteration.  All
  algorithms use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank2}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank3}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing Pick Best with previous results.  All algorithms use
  the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank4}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.  All algorithms use the
  Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

With Griewank, the premature convergence problems inherent in picking the best
child are exacerbated when speculating several iterations ahead.  When Many
Iterations finds the optimum, it finds it quicker than any other method we
tried, on average four times faster than standard PSO.  However, it also gets
stuck and fails to find the optimum more than any other method.  The results
are shown in \fig{griewank-manyiters}.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank5}
  \caption{Function Griewank with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.  All algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{griewank6}
  \caption{Function Griewank with a swarm that uses 800 processors per round of
  evaluations.  All algorithms use the Ring topology.}
  \label{fig:griewank-800}
\end{figure}

@@

Bohachevsky is a unimodal function best optimized with a Complete swarm.  It is
similar to Griewank in that there is a global optimum with a value of 0, and
the swarm sometimes finds it and sometimes does not.  Since we have already
demonstrated the convergence properties of the various algorithms with
Griewank, for Bohachevsky we merely present the best results.  In
\fig{bohachevsky} we can see that all of the speculative algorithms found the
optimum much quicker than PSO with a Random topology.  However, SEPSO was
slower than PSO Complete and got stuck around 15\% of the time.  Using Pick
Best, Pick Best Pruned, and Many Iterations all outperformed PSO Complete,
though Many Iterations only almost reached a 100\% success rate.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with a swarm that uses 800 processors per round
  of evaluations.  All of the speculative algorithms use the Random topology.}
  \label{fig:bohachevsky}
\end{figure}
